Sparse Function-space Representation of Neural Networks

### 2023-05-30 workshop/main.tex
smalls
Edits

\begin{abstract}
Deep neural networks are known to lack uncertainty estimates, struggle to incorporate new data, and suffer from catastrophic forgetting. We present a method that mitigates these issues by converting neural networks from weight-space to a low-rank function-space representation, via the so-called dual parameters. In contrast to previous work, our sparse representation captures the joint distribution over the entire data set, rather than only over a subset. This offers a compact and principled way of capturing uncertainty and enables us to incorporate new data without retraining whilst retaining predictive performance. We demonstrate the proposed approach for quantifying uncertainty in supervised learning and maintaining an expressive functional representation for sequential learning.\looseness-1
\end{abstract}

\section{Introduction}
%
Deep learning \cite{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data? These questions become central when applied to sequential learning paradigms, such as continual learning (CL, \citep{parisi2019continual, de2021continual}) and reinforcement learning (RL, \cite{sutton2018reinforcement}). In CL, access to the previous data is lost, and then the challenge is retaining a compact representation of the problem to alleviate forgetting over the life-long learning horizon~\cite{mccloskey1989catastrophic}. Similarly, in RL, the model must adapt to environmental observations through exploration, while leveraging prediction uncertainties to assess potential future paths.\looseness-1

@@

%Current state of affairs 
Recent techniques (\eg, \cite{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable}) apply a Laplace-GGN approximation to convert trained neural networks into Bayesian neural networks, that can provide uncertainty without sacrificing additional resources to training \cite{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \cite{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing its behaviour \cite{cho2009kernel,meronen2020stationary}, performing probabilistic inference \cite{khan2019approximate}, and quantifying uncertainty in neural networks \cite{foong2019between}. These methods rely on the linearization of the neural network and the resultant neural tangent kernel (NTK, \cite{jacot2018neural}). The neural network is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates. They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits `dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \cite{immer2021scalable},
this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. 
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

%
We evaluate \our on eight UCI benchmarks~\cite{UCI}, a variety of binary and multi-class classification tasks with different data set sizes.
We train a two-layer MLP for each of the classification tasks and follow the experiment set up in \cite{immer2021improving}.
%by using the same hyperparameters, performing a hyperparameter search over the prior precision $\delta$,
% We follow \cite{immer2021improving} by using the same hyperparameters, performing a hyperparameter search over the prior precision $\delta$,
% and run the experiment over $10$ random splits.
% That is, after training we construct the \our dual and use the resulting model for uncertainty quantification.
% for the neural network training as in the UCI experiments of
\cref{tbl:uci} (left) shows that \our with $M=256$ either matches or outperforms the predictive performance of the NN MAP, mean-field VI, Bayesian NN, and GLM
predictions %in terms of negative log-predictive density (NLPD).
(baselines from \cite{immer2021improving}).
That is, we match predictive performance despite being sparse.
%
In \cref{tbl:uci} (right) we compare to the subset GP method from \cite{immer2021improving} whilst using only $M=32$ inducing points.
It shows that \our is able to summarize the full data set more effectively than the GP subset method as it maintains predictive performance
whilst using fewer inducing points.
\cref{fig:uci} further shows that as the number of inducing points is lowered $M=512,\ldots, 16$, \our is able to maintain a much better NLPD than the GP subset.
These results demonstrate \our's sparse representation captures information from the entire data set and as a result provides good uncertainty estimates.
We provide further details of our experiments in \cref{app:uci}.
% as well as further results with varying number of inducing poitns ,64,128,256$
% We provide full details of our experiments as well as further experiments with varying number of inducing poitns and further results using $M=16,32,64,128,256$ in \cref{app:uci}.
% benefits of \our in summarizing the full data distribution onto a small set of inducing
% points over just picking a random subset.

\section{Discussion and conclusion}
\label{sec:conclusion}
%
In this paper, we have introduced \our, a novel approach for representing neural networks in sparse function space, exploiting the dual parameters for an efficient low-rank approximation that accommodates information from the entire data distribution. Our method offers a powerful mechanism for capturing predictive uncertainty, updating the model with new data without retraining, and providing a compact representation suitable for continual learning. These aspects were demonstrated in a wide range of problems, data sets, and learning contexts. We showcased our method's ability to capture uncertainty in UCI classification tasks (\cref{sec:uci}), demonstrated robustness on image data sets (\cref{,sec:image}), established its potential for continual learning (\cref{sec:cl-exp}), and finally, verified its applicability in reinforcement learning scenarios (\cref{sec:rl-exp}).

>>>>

\begin{abstract}
Deep neural networks are known to lack uncertainty estimates, struggle to incorporate new data, and suffer from catastrophic forgetting. We present a method that mitigates these issues by converting neural networks from weight-space to a low-rank function-space representation, via the so-called dual parameters. In contrast to previous work, our sparse representation captures the joint distribution over the entire data set, rather than only over a subset. This offers a compact and principled way of capturing uncertainty and enables us to incorporate new data without retraining whilst retaining predictive performance. We provide proof-of-concept demonstrations with the proposed approach for quantifying uncertainty in supervised learning on UCI benchamark tasks.
\end{abstract}

\section{Introduction}
%
Deep learning \cite{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data? These questions become central when applied to sequential learning paradigms, such as continual learning \citep[CL,][]{parisi2019continual, de2021continual} and reinforcement learning  \citep[RL,][]{sutton2018reinforcement}. In CL, access to the previous data is lost, and then the challenge is retaining a compact representation of the problem to alleviate forgetting over the life-long learning horizon~\citep{mccloskey1989catastrophic}. Similarly, in RL, the model must adapt to environmental observations through exploration, while leveraging prediction uncertainties to assess potential future paths.\looseness-1

@@

%Current state of affairs 
Recent techniques  \citep[\eg,][]{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable} apply a Laplace-GGN approximation to convert trained neural networks into Bayesian neural networks, that can provide uncertainty without sacrificing additional resources to training \cite{foong2019between}. Furthermore, the resultant weight-space posterior can be converted to the function-space as shown in \citet{khan2019approximate, immer2021improving}. The function-space representation allows for a principled mathematical approach for analyzing its behaviour \citep{cho2009kernel,meronen2020stationary}, performing probabilistic inference \citep{khan2019approximate}, and quantifying uncertainty in neural networks \citep{foong2019between}. These methods rely on the linearization of the neural network and the resultant neural tangent kernel \citep[NTK,][]{jacot2018neural}. The neural network is characterized in function-space by its first two moment functions, a mean function and covariance function (or kernel)---defining a Gaussian process \citep[GP,][]{rasmussen2006gaussian}. GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates. They serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}.
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits `dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \citep{immer2021scalable}, this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs. 
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. 
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network; see \cref{fig:teaser} for an example. %Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach for conditioning on new data without needing to retrain the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

%
We provide a proof-of-concept example by evaluating \our on eight UCI benchmarks~\cite{UCI}, a variety of binary and multi-class classification tasks with different data set sizes. 
We train a two-layer MLP for each of the classification tasks and follow the experiment set up in \citet{immer2021improving}.
%by using the same hyperparameters, performing a hyperparameter search over the prior precision $\delta$,
% We follow \cite{immer2021improving} by using the same hyperparameters, performing a hyperparameter search over the prior precision $\delta$,
% and run the experiment over $10$ random splits.
% That is, after training we construct the \our dual and use the resulting model for uncertainty quantification.
% for the neural network training as in the UCI experiments of
\cref{tbl:uci} (left) shows that \our with $M=256$ either matches or outperforms the predictive performance of the NN MAP, mean-field VI, Bayesian NN, and GLM
predictions %in terms of negative log-predictive density (NLPD).
\citep[baselines from][]{immer2021improving}.
That is, we match predictive performance despite being sparse.
%
In \cref{tbl:uci} (right) we compare to the subset GP method from \cite{immer2021improving} whilst using only $M=32$ inducing points.
It shows that \our is able to summarize the full data set more effectively than the GP subset method as it maintains predictive performance
whilst using fewer inducing points.
\cref{fig:uci} further shows that as the number of inducing points is lowered $M=512,\ldots, 16$, \our is able to maintain a much better NLPD than the GP subset.
These results demonstrate \our's sparse representation captures information from the entire data set and as a result provides good uncertainty estimates. See \citet{immer2021improving} for further details on this particular benchmark setup.
%We provide further details of our experiments in \cref{app:uci}.
% as well as further results with varying number of inducing poitns ,64,128,256$
% We provide full details of our experiments as well as further experiments with varying number of inducing poitns and further results using $M=16,32,64,128,256$ in \cref{app:uci}.
% benefits of \our in summarizing the full data distribution onto a small set of inducing
% points over just picking a random subset.

\section{Discussion and conclusion}
\label{sec:conclusion}
%
In this paper, we have introduced \our, a novel approach for representing neural networks in sparse function space, exploiting the dual parameters for an efficient low-rank approximation that accommodates information from the entire data distribution. Our method offers a powerful mechanism for capturing predictive uncertainty, opens up new possible avenues for updating the model with new data without retraining, and providing a compact representation suitable for continual learning. In this workshop paper, we showcased our method's ability to capture uncertainty in UCI benchmark classification tasks (\cref{sec:uci}). The appealing properties of \our should make it a good addition to the tool set of modelling in sequential learning settings, such as continual and reinforcement learning. 
%These aspects were demonstrated in a wide range of problems, data sets, and learning contexts. We showcased our method's ability to capture uncertainty in UCI classification tasks (\cref{sec:uci}), demonstrated robustness on image data sets (\cref{,sec:image}), established its potential for continual learning (\cref{sec:cl-exp}), and finally, verified its applicability in reinforcement learning scenarios (\cref{sec:rl-exp}).
